---
title: "Interactive Heatmap"
author: "Nicholas Louque"
date: "`r Sys.Date()`"
output: html_document
---



###############################################################################
#   This function reads:
#     1) an expression matrix (genes x cells) from a CSV file, and
#     2) per-cell metadata
#     - optionally filters cells and/or genes via provided selection CSVs,
#     - groups cells into hierarchical "paths" using metadata columns
#       (e.g., class -> subclass -> cluster),
#     - computes the average expression per gene per group,
#     - optionally normalizes/scales the resulting group-average matrix,
#     - generates an interactive heatmap with:
#         * clustering (rows + optionally columns),
#         * multiple metadata annotation bars across columns,
#         * a cleaned legend so viewers see actual labels (not hex codes),
#         * hover text that reports annotation variable/value cleanly.
#
#   You must provide TWO CSV files:
#
#   (A) expression_matrix CSV  (REQUIRED)
#       - Path passed as `expression_matrix`
#       - Format: genes as rows, cells as columns.
#       - FIRST column is gene names (row names). Example:
#           gene,cellA,cellB,cellC
#           MS4A1,0,1.2,0.3
#           LST1,4.5,0.1,2.2
#       - This code loads it with:
#           read.csv(expression_matrix, row.names = 1, stringsAsFactors = FALSE)
#       - So the first column must be gene IDs and becomes rownames(expression_data).
#
#   (B) metadata CSV  (REQUIRED)
#       - Path passed as `metadata`
#       - Format: FIRST column is cell IDs (must match expression column names).
#         The remaining columns are hierarchical grouping labels, ordered from
#         broad -> fine (left -> right).
#       - Example:
#           cell_id,class,subclass,cluster
#           cellA,Immune,Tcell,Tcell_1
#           cellB,Immune,Bcell,Bcell_2
#           cellC,Glia,Microglia,Micro_3
#       - IMPORTANT:
#         * The first column is treated as the cell ID column even if it’s named
#           something else. If it has no name, the function renames it to "cell_id".
#         * Any metadata columns after the first are treated as hierarchy levels.
#
# OPTIONAL INPUT FILES 
#   1) gene_selection (CSV, 1 column, no header)
#       - Path passed as `gene_selection`
#       - A plain list of genes to keep (one per row).
#       - Example file (selected_genes.csv):
#           MS4A1
#           LST1
#           IL7R
#
#   2) cell_selection (CSV, 1 column, no header)
#       - Path passed as `cell_selection`
#       - A plain list of cell IDs to keep (one per row).
#       - Example file:
#           cellA
#           cellC
#
#   3) group_by (CSV, header optional; function reads first cell of first column)
#       - Path passed as `group_by`
#       - If provided, this file tells the function which metadata column is the
#         "finest level" used to build grouping paths (how far down the hierarchy).
#       - Example file (group_by_level.csv):
#           level
#           cluster
#
#   4) group_order (CSV, any shape; function flattens values)
#       - Path passed as `group_order`
#       - If provided, it controls which metadata columns are used (and in what
#         order) for the hierarchy and annotation bars.
#       - Example file:
#           class
#           subclass
#           cluster
#
#
# NORMALIZATION OPTIONS (argument: normalization)
#   - "none"        : no transform
#   - "log2"        : log2(expr + 1) applied BEFORE averaging (on cell values)
#   - "scale"       : z-score per gene AFTER averaging (row-wise scale)
#   - "log2_scale"  : log2(expr + 1) then z-score per gene after averaging
#   - "max"         : divide each gene row by its max value after averaging
# Notes:
#   * Scaling is performed on the gene rows: t(scale(t(expr_matrix))).
#   * NA/Inf are replaced with 0 before plotting.
#
# COLOR OPTIONS (argument: colorscale)
#   - "viridis" (default), "plasma", "Blues", "Reds"
#   Used for the heatmap intensity scale (expression values).
#
# ANNOTATION BARS + LEGEND
#   For each hierarchy level used (grouping_cols), the function creates an
#   annotation bar across columns using distinct colors per category value.
#   It then post-processes the plotly object to:
#     - replace hex-only legend entries with readable labels,
#     - ensure each annotation value appears once in the legend,
#     - move the legend to the right with extra right margin.
#
# CLUSTERING / DENDROGRAM BEHAVIOR
#   - If group_order is provided: dendrogram="row" and column dendrogram hidden
#   - Otherwise: dendrogram="both" and both dendrograms shown
#
# PARAMETERS
#   MoussaMap(
#     expression_matrix = "DS1.csv",     # REQUIRED path to expression CSV
#     metadata          = "hierarchy.csv",# REQUIRED path to metadata CSV
#     gene_selection    = NULL,          # OPTIONAL path to gene list CSV
#     cell_selection    = NULL,          # OPTIONAL path to cell list CSV
#     group_by          = NULL,          # OPTIONAL path to file containing finest level name
#     group_order       = NULL,          # OPTIONAL path to file containing metadata col order
#     group_by_level    = NULL,          # OPTIONAL direct finest-level name (overridden by group_by file)
#     colorscale        = "viridis",     # OPTIONAL heatmap color scale
#     normalization     = "none",        # OPTIONAL normalization mode
#     width             = 1000,          # OPTIONAL plot width (pixels)
#     height            = 800            # OPTIONAL plot height (pixels)
#   )
#
# RETURN VALUE
#   A list:
#     - $plot  : plotly/heatmaply object (interactive; render in RMarkdown/HTML)
#     - $error : (only on failure) error string
#   Typical usage:
#     result <- MoussaMap(...)
#     result$plot
#
#
###############################################################################



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Install and load packages
required_packages <- c(
  "shiny", "shinydashboard", "heatmaply", "dplyr", "DT", "tidyr", "tibble",
  "RColorBrewer", "viridis", "randomcoloR", "stringr", "plotly", "ggdendro",
  "Seurat"
)

for(pkg in required_packages) {
  if(!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```
```{r heatmap-function}
# Helper: Select VariableFeatures AFTER averaging (genes x groups)
select_variable_features_after_averaging <- function(expr_matrix,
                                                    nfeatures = 2000,
                                                    selection.method = "vst") {
  expr_matrix <- as.matrix(expr_matrix)

  # Treat "groups" as Seurat "cells"
  seu <- Seurat::CreateSeuratObject(counts = expr_matrix)

  # Normalization is required for VST/HVG workflow
  seu <- Seurat::NormalizeData(seu, verbose = FALSE)
  seu <- Seurat::FindVariableFeatures(
    seu,
    selection.method = selection.method,
    nfeatures = min(nfeatures, nrow(expr_matrix)),
    verbose = FALSE
  )

  Seurat::VariableFeatures(seu)
}

# Main heatmap creation function
MoussaMap <- function(expression_matrix,
                          metadata,
                          gene_selection = NULL,
                          cell_selection = NULL,
                          group_by = NULL,
                          group_order = NULL,
                          group_by_level = NULL,
                          top_var_features = NULL,   # NEW (e.g., 50/200/2000). NULL disables
                          varfeat_method = "vst",    # NEW ("vst" default)
                          colorscale = "viridis",
                          normalization = "none",
                          width = 1000,
                          height = 800) {

  tryCatch({
    # Read expression and metadata files
    if(!file.exists(expression_matrix)) stop("Expression matrix CSV not found: ", expression_matrix)
    if(!file.exists(metadata)) stop("Metadata CSV not found: ", metadata)
    
    expr_temp <- read.csv(expression_matrix, stringsAsFactors = FALSE, check.names = FALSE)
    gene_names <- make.unique(trimws(as.character(expr_temp[, 1])), sep = "_")
    rownames(expr_temp) <- gene_names
    expression_data <- expr_temp[, -1, drop = FALSE]
    metadata_data   <- read.csv(metadata, stringsAsFactors = FALSE, check.names = FALSE)
    
    # Set cell ID column name
    if (is.null(names(metadata_data)[1]) || names(metadata_data)[1] == "") {
      names(metadata_data)[1] <- "cell_id"
    }
    id_col <- names(metadata_data)[1]
    metadata_cols <- names(metadata_data)[-1]
    
    # Find common cells between datasets
    common_cells <- intersect(colnames(expression_data), metadata_data[[id_col]])
    if(length(common_cells) == 0) stop("No matching cells found between expression matrix and metadata.")

    # Filter cells if selection provided
    if (!is.null(cell_selection) && file.exists(cell_selection)) {
      cell_list <- unique(trimws(read.csv(cell_selection, header = FALSE, stringsAsFactors = FALSE)[[1]]))
      cell_list <- cell_list[cell_list != ""]
      if (length(cell_list) > 0) {
        common_cells <- intersect(common_cells, cell_list)
      } else {
        warning("cell_selection was empty — ignored.")
      }
    }

    # Remove duplicates and align data
    expression_data <- expression_data[, !duplicated(colnames(expression_data))]
    metadata_data   <- metadata_data[!duplicated(metadata_data[[id_col]]), ]
    expression_data <- expression_data[, common_cells, drop = FALSE]
    metadata_data   <- metadata_data[metadata_data[[id_col]] %in% common_cells, , drop = FALSE]
    idx <- match(common_cells, metadata_data[[id_col]])
    metadata_data <- metadata_data[idx, , drop = FALSE]

    # Filter genes if selection provided
    if (!is.null(gene_selection) && file.exists(gene_selection)) {
      gene_list <- unique(trimws(read.csv(gene_selection, header = FALSE, stringsAsFactors = FALSE)[[1]]))
      gene_list <- gene_list[gene_list != ""]
      if (length(gene_list) > 0) {
        expression_data <- expression_data[rownames(expression_data) %in% gene_list, , drop = FALSE]
      } else {
        warning("gene_selection was empty — ignored.")
      }
    }

    # Apply log2 normalization if needed (cell-level) BEFORE averaging
    if (normalization == "log2" || normalization == "log2_scale") {
      expression_data <- log2(expression_data + 1)
    }
    
    # Read grouping level if provided
    if (!is.null(group_by) && file.exists(group_by)) {
      group_info <- read.csv(group_by, header = TRUE, stringsAsFactors = FALSE)
      chosen_level <- trimws(as.character(group_info[[1]][1]))
      if (!is.na(chosen_level) && chosen_level != "") {
        group_by_level <- chosen_level
      } else {
        warning("group_by empty — using default finest level.")
        group_by_level <- NULL
      }
    }
    
    # Read custom column order
    if (!is.null(group_order) && file.exists(group_order)) {
      order_info  <- read.csv(group_order, header = TRUE, stringsAsFactors = FALSE)
      group_order <- unique(na.omit(trimws(unlist(order_info))))
      group_order <- group_order[group_order != ""]

      if (length(group_order) > 0) {
        if ("class" %in% group_order && "class" %in% names(metadata_data)) {
          group_by_level <- "class"
        }
        metadata_cols <- intersect(group_order, metadata_cols)
      } else {
        warning("group_order was empty or invalid.")
        group_order <- NULL
      }
    }
    
    # Determine hierarchy levels to use
    metadata_cols <- trimws(metadata_cols)
    
    if (!is.null(group_by_level) && group_by_level %in% metadata_cols) {
      finest_level <- group_by_level
    } else {
      finest_level <- tail(metadata_cols, 1)
    }
    
    finest_index <- match(finest_level, metadata_cols)
    grouping_cols <- metadata_cols[1:finest_index]
    
    # Create full hierarchical path
    metadata_data$full_path <- apply(
      metadata_data[, grouping_cols, drop = FALSE], 1,
      function(x) paste(x, collapse = "_")
    )
    
    if (is.null(group_by_level)) {
      group_by_level <- tail(metadata_cols, 1)
    }
    
    # Assign cells to groups
    cell_assignments <- setNames(metadata_data$full_path, metadata_data[[id_col]])
    
    # Combine data for averaging
    combined_data <- expression_data %>%
      tibble::rownames_to_column("gene") %>%
      tidyr::pivot_longer(-gene, names_to = "cell", values_to = "expression") %>%
      dplyr::mutate(cell_group = cell_assignments[cell]) %>%
      dplyr::filter(!is.na(cell_group))
    
    # Calculate group averages
    cell_type_averages <- combined_data %>%
      dplyr::group_by(gene, cell_group) %>%
      dplyr::summarise(avg_expression = mean(expression, na.rm = TRUE), .groups = 'drop')

    # Convert to matrix format (genes x groups)
    expr_matrix <- cell_type_averages %>%
      tidyr::pivot_wider(names_from = cell_group, values_from = avg_expression, values_fill = 0) %>%
      tibble::column_to_rownames("gene") %>%
      as.matrix()

    # Clean invalid values
    if (any(is.na(expr_matrix)) || any(is.infinite(expr_matrix))) {
      expr_matrix[is.na(expr_matrix)] <- 0
      expr_matrix[is.infinite(expr_matrix)] <- 0
    }

    # ============================
    # NEW: VariableFeatures AFTER averaging (before scaling/max)
    # ============================
    if (!is.null(top_var_features) && is.numeric(top_var_features) && top_var_features > 0) {
      var_genes <- select_variable_features_after_averaging(
        expr_matrix = expr_matrix,
        nfeatures = top_var_features,
        selection.method = varfeat_method
      )
      var_genes <- intersect(var_genes, rownames(expr_matrix))
      if (length(var_genes) == 0) {
        warning("top_var_features requested, but no VariableFeatures matched rownames(expr_matrix). Skipping.")
      } else {
        expr_matrix <- expr_matrix[var_genes, , drop = FALSE]
      }
    }
    
    # Apply scaling normalization (AFTER selecting features)
    if (normalization == "scale" || normalization == "log2_scale") {
      expr_matrix <- t(scale(t(expr_matrix)))
    } else if (normalization == "max") {
      max_vals <- apply(expr_matrix, 1, max, na.rm = TRUE)
      max_vals[max_vals == 0] <- 1
      expr_matrix <- expr_matrix / max_vals
    }

    # Build hierarchy data frame
    col_groups <- colnames(expr_matrix)
    hierarchy_df <- data.frame(full_path = col_groups, stringsAsFactors = FALSE)
    
    for (i in seq_along(grouping_cols)) {
      hierarchy_df[[grouping_cols[i]]] <- sapply(strsplit(col_groups, "_"), function(x) {
        if (length(x) >= i) x[i] else NA
      })
    }
    
    # Generate colors for each level.
    # Across levels: each level gets a completely different hue family so bars
    # are immediately distinguishable from one another.
    # Within each level: colors are ordered dark/light/medium to maximise
    # perceptual contrast between adjacent categories so every color pops.
    col_side_colors <- list()
    legend_data <- list()

    hue_families <- list(
      # Blues: navy -> sky -> royal -> pale -> cobalt -> powder -> mid -> ice
      c("#08306b","#6baed6","#2171b5","#c6dbef","#08519c","#9ecae1","#4292c6","#deebf7"),
      # Reds: burgundy -> salmon -> crimson -> blush -> scarlet -> rose -> brick -> pink
      c("#67000d","#fb6a4a","#a50f15","#fcbba1","#cb181d","#fc9272","#ef3b2c","#fee0d2"),
      # Greens: forest -> mint -> emerald -> honeydew -> hunter -> sage -> fern -> pale
      c("#00441b","#74c476","#238b45","#c7e9c0","#006d2c","#a1d99b","#41ab5d","#e5f5e0"),
      # Purples: indigo -> lavender -> violet -> lilac -> plum -> periwinkle -> grape -> pale
      c("#3f007d","#9e9ac8","#54278f","#dadaeb","#6a51a3","#bcbddc","#807dba","#f2f0f7"),
      # Oranges: rust -> peach -> burnt -> apricot -> tangerine -> buff -> amber -> cream
      c("#7f2704","#fd8d3c","#a63603","#fdd0a2","#d94801","#fdae6b","#f16913","#fee6ce"),
      # Teals: deep -> aqua -> pine -> mist -> teal -> seafoam -> cyan -> pale
      c("#003c30","#80cdc1","#01665e","#c7eae5","#35978f","#b2e2e2","#5ab4ac","#edf8fb")
    )

    for (level_idx in seq_along(grouping_cols)) {
      level_name <- grouping_cols[level_idx]
      level_values <- hierarchy_df[[level_name]]
      unique_vals <- unique(level_values[!is.na(level_values)])
      n_unique <- length(unique_vals)

      pal <- hue_families[[(level_idx - 1) %% length(hue_families) + 1]]
      colors <- if (n_unique <= length(pal)) {
        pal[1:n_unique]
      } else {
        colorRampPalette(pal)(n_unique)
      }

      color_map <- setNames(colors, unique_vals)
      color_vals <- paste(level_values, color_map[level_values], sep = "||")
      col_side_colors[[level_name]] <- color_vals
      legend_data[[level_name]] <- data.frame(label = unique_vals, color = colors)
    }
    
    # Prepare annotation colors
    col_side_colors <- as.data.frame(col_side_colors, row.names = col_groups)
    col_side_colors <- col_side_colors[, rev(colnames(col_side_colors)), drop = FALSE]
    
    # Select color palette
    color_func <- switch(colorscale,
                         viridis = viridis::viridis,
                         plasma  = viridis::plasma,
                         Blues   = function(n) RColorBrewer::brewer.pal(n, "Blues"),
                         Reds    = function(n) RColorBrewer::brewer.pal(n, "Reds"),
                         viridis::viridis)
    
    # Configure dendrogram display
    if (!is.null(group_order)) {
      dendro_mode <- "row"
      show_dend   <- c(TRUE, FALSE)
    } else {
      dendro_mode <- "both"
      show_dend   <- c(TRUE, TRUE)
    }
    
    # Build plot title
    norm_label <- switch(normalization,
                        "log2" = " (log2 normalized)",
                        "scale" = " (scaled)",
                        "log2_scale" = " (log2 + scaled)",
                        "max" = " (max scaled)",
                        "none" = " (no normalization)",
                        "")

    # Create heatmap plot
    col_side_colors[] <- lapply(col_side_colors, function(x) as.character(x))
    heatmap_plot <- heatmaply::heatmaply(
      expr_matrix,
      colors = color_func(256),
      dendrogram = dendro_mode,
      show_dendrogram = show_dend,
      col_side_colors = col_side_colors,
      main = paste0("Expression Heatmap — grouped by ", finest_level, norm_label),
      xlab = "Groups", ylab = "Genes",
      fontsize_row = 8, fontsize_col = 6,
      hclust_method = "ward.D2", dist_method = "euclidean",
      width = width, height = height
    )

    annotation_order <- colnames(col_side_colors)
    plt <- heatmap_plot

    # Create legend entries
    legend_entries <- list()
    seen <- character()

    for (level in annotation_order) {
      for (tr in plt$x$data) {
        if (!is.null(tr$fillcolor) && !is.null(tr$text)) {
          variable <- sub(".*variable: ([^<]+).*", "\\1", tr$text)
          value    <- sub(".*value: ([^<]+).*", "\\1", tr$text)

          if (variable != level) next

          key <- paste(variable, value, tr$fillcolor)

          if (!(key %in% seen)) {
            seen <- c(seen, key)
            legend_entries[[length(legend_entries) + 1]] <- list(
              type = "scatter",
              mode = "markers",
              x = NA,
              y = NA,
              marker = list(size = 10, color = tr$fillcolor),
              name = paste(variable, "=", value),
              legendgroup = variable,
              showlegend = TRUE,
              hoverinfo = "skip"
            )
          }
        }
      }
    }

    # Hide hex color legends
    for (i in seq_along(plt$x$data)) {
      tr <- plt$x$data[[i]]
      if (!is.null(tr$name) && grepl("^#[0-9A-Fa-f]{6}$", tr$name)) {
        plt$x$data[[i]]$showlegend <- FALSE
      }
    }

    plt$x$data <- c(plt$x$data, legend_entries)
    heatmap_plot <- plt

    # Update hover and legend labels
    new_data <- list()
    seen_legend <- character()

    for (tr in plt$x$data) {
      if (!is.null(tr$fillcolor) && !is.null(tr$text) && !is.null(tr$x)) {
        level <- sub(".*variable: ([^<]+).*", "\\1", tr$text)
        if (!level %in% grouping_cols) {
          new_data[[length(new_data) + 1]] <- tr
          next
        }

        group_name <- tr$x[1]
        idx <- match(group_name, hierarchy_df$full_path)
        if (is.na(idx)) {
          new_data[[length(new_data) + 1]] <- tr
          next
        }

        meta_value <- hierarchy_df[[level]][idx]

        tr$text <- paste0("variable: ", level, "<br>", "value: ", meta_value)

        legend_key <- paste(level, meta_value)
        tr$name <- meta_value
        tr$legendgroup <- level
        
        if (!(legend_key %in% seen_legend)) {
          tr$showlegend <- TRUE
          seen_legend <- c(seen_legend, legend_key)
        } else {
          tr$showlegend <- FALSE
        }

        new_data[[length(new_data) + 1]] <- tr
      } else {
        new_data[[length(new_data) + 1]] <- tr
      }
    }

    plt$x$data <- new_data
    heatmap_plot <- plt

    # Clean legend labels
    plt <- heatmap_plot

    for (i in seq_along(plt$x$data)) {
      tr <- plt$x$data[[i]]
      if (isTRUE(tr$showlegend) && !is.null(tr$name)) {
        if (grepl("\\|\\|#", tr$name)) {
          cleaned <- sub(".*=\\s*", "", tr$name)
          cleaned <- sub("\\|\\|#.*$", "", cleaned)
          level <- tr$legendgroup
          plt$x$data[[i]]$name <- paste0(level, ": ", cleaned)
        }
      }
    }

    heatmap_plot <- plt

    # Configure legend position
    heatmap_plot$x$layout$legend <- list(
      orientation = "v",
      x = 1.22,
      y = 1,
      xanchor = "left",
      yanchor = "top",
      font = list(size = 12)
    )

    # Set plot margins
    heatmap_plot$x$layout$margin <- list(
      l = 80,
      r = 260,
      b = 80,
      t = 80
    )

    heatmap_plot$x$layout$showlegend <- TRUE

    legend_traces <- list()
    heatmap_plot$x$data <- c(heatmap_plot$x$data, legend_traces)

    # Hide annotation rectangle legends
    plt <- heatmap_plot

    for (i in seq_along(plt$x$data)) {
      tr <- plt$x$data[[i]]
      if (!is.null(tr$x) && !all(is.na(tr$x))) {
        plt$x$data[[i]]$showlegend <- FALSE
      }
    }

    heatmap_plot <- plt

    # Return plot object
    return(list(plot = heatmap_plot))
    
  }, error = function(e) {
    traceback()
    list(plot = NULL, error = paste("Error:", e$message))
  })
}
```

```{r}
# Test with Suryaveer's files
result <- MoussaMap(
  expression_matrix = "expr_matrix_liver.csv",
  metadata = "cell_id_class_subclass_cluster.csv",
  normalization = "none"
)
result$plot
```
```{r}
# Log2 transform plus scaling
result_log_scale <- MoussaMap(
  expression_matrix = "DS1.csv",
  metadata = "hierarchy10.csv",
  normalization = "log2_scale",
  colorscale = "viridis",
)
result_log_scale$plot
```
```{r}
# Max normalization
result_max <- MoussaMap(
  expression_matrix = "DS1.csv",
  metadata = "hierarchy10.csv",
 
)
result_max$plot
```
```{r}
# Z-score scaling
result_scale <- MoussaMap(
  expression_matrix = "DS1.csv",
  metadata = "hierarchy10.csv",
  normalization = "scale",
  group_by = "group_by_level.csv"
)
result_scale$plot
```
```{r}
# Log2 transform only
result_log <- MoussaMap(
  expression_matrix = "DS1.csv",
  metadata = "hierarchy10.csv",
  normalization = "log2"
)
result_log$plot
```
